Funding programmes for early career researchers marginally improves their collaboration independence, but not cognitive independence

### 2023-11-01 article.Rmd
changes in structure of the control groups to better match "funded" and "unfunded" terms

## Intro 1: focus on the bigger picture, sociology and policy perspective

Researchers’ independence from their PhD advisor is one of the criteria often used for assessing researchers for hiring, tenure and promotion (@vanarensbergenDifferentViewsScholarly2014; @moherAssessingScientistsHiring2018), constituting one factor of research excellence.
The promise of independence is that researchers will “challenge the existing common views and open up new lines of research” (@vandenbesselaarMeasuringResearcherIndependence2019, p.2).

Early career researcher independence is central to the long-lasting debate about how science progresses - whether it is indeed guided mostly by the truth and empirical evidence, or whether more social and demographic factors might play an important role. @hull1978 found that even more senior researchers readily updated their beliefs with the new evidence, supporting the former, though many (famously including Max Planck) observed that senior researchers were more fixed in their beliefs and new ideas were mostly brought up by the incoming early career researchers giving rise to the “science moves forward one funeral at the time” proverb. More recently, @azoulay2010 provided further support for the latter by investigating the effects of unexpected deaths of prominent scientists on their research fields and their collaborators' careers. They found a decrease in the productivity of junior collaborators and an increase in published work by young newcomers to the field rewarded by an increased citation rate. That suggests the sudden departures of prominent scientists gave room for more disruptive changes in the field and for bringing new ideas and perspectives while harming the status quo holders who worked with the prominent scientists on their research agendas - a “Goliath effect”, as coined by @wang2021, p.42).

/// here I can potentially talk a bit more about Inheritance of research agendas - how often does that happen, etc..///

Insofar as the latter is true, in order to accelerate scientific progress, we might want to tackle this trend by supporting early career researchers to gain more independence from their PhD advisors. Younger researchers are already more likely to come up with new topics and ideas, as found by @packalen2019 by looking at the usage of new words in research papers. However, they might still feel various kinds of pressures to stick with their mentor’s research agendas and lack of opportunities to become more independent. In recent decades, there has been a worry that researchers’ independence is decreasing, illustrated by an increase in age at which US biomedical researchers received their first grant (NAS, 2005; @danielsGenerationRiskYoung2015), as well as direct measures of new researchers becoming more dependent on their PhD advisors across almost all disciplines in Slovenia (@rojkoScientificPerformanceResearch2022, table 6 on p.9).

Perhaps to tackle this trend, many grant agencies decided to start awarding grants for early career researchers, usually up to 8 years after finishing their PhD. Such programmes have often explicitly stated goals of helping researchers become more independent, set up their own research group and open up new lines of research. The examples of such programmes could be Starting Grants by European Research Council (@StartingGrantERC), Junior project and Junior Star by Czech Science Foundation (2020), Walter N. Benjamin Programme by DFG (@WalterBenjaminProgramme), Innovation Research Incentives Scheme by Netherlands Organization of Scientific Research (NWO) (2021).

Moreover, stimulating independence and supporting researchers to open up new lines of research is also a worthy policy goal regardless of researchers’ age. @rzhetsky2015 found that biomedical researchers are in practice often choosing less exploratory strategies for topic choice than would be socially optimal if we wanted to explore the cognitive map of science (or in this case chemicals) as fast as possible, and their choices become more conservative over time. They suggest that increased interdisciplinarity, risk-taking and the publication of experimental failures would substantially improve the speed of discovery.

//// factors influencing independence////
There are few known factors influencing the level of research independence. One of them is study discipline. Based on some early empirical studies (@rojkoScientificPerformanceResearch2022), it seems that researchers in disciplines such as philosophy, psychology; religion, theology; social sciences; the arts, recreation, entertainment, sport; language, linguistics, literature; and geography, biography, history experience significantly higher levels of independence on their PhD advisor than researchers in disciplines such as science and knowledge, organization, computer science, information, documentation, librarianship, institutions, publications; mathematics, natural sciences; and applied sciences, medicine, technology.

Interestingly, there is also some evidence that the propensity to cognitive independence might be inherited and passed down from generation to generation in mentor-mentee relationships @yoshioka-kobayashi2021 - perhaps it is not only research agendas that students inherit from their advisors but also the tendency to deviate from those research agendas. This might also suggest that to increase junior researchers’ independence, we might want to teach their advisors to be more independent, in order to have a ripple effect down the line.

Finally, there is also some preliminary evidence that higher independence modestly correlates with the number of citations acquired (@rojkoScientificPerformanceResearch2022), but others argue that even the research systems with very low independence can create a lot of scientific impact, and thus the relationships between independence and impact might not be as clear (@glaserIndependenceResearchReview2022)

Although researcher independence seems to be an important feature that is acknowledged by many actors in the research systems and that research funders try to stimulate, there hasn’t been much effort focused on quantifying this phenomenon. Some studies try to measure independence in a simple way, as a ratio of publications the given researcher produced without their supervisor divided by the total number of publications of the given researchers (e.g. @rojkoScientificPerformanceResearch2022). However, such an indicator could be misleading in the world of team science, and also only focuses on collaborations, not tracking whether a given researcher becomes independent in their thinking and research topics. [@vandenbesselaarMeasuringResearcherIndependence2019] therefore proposed a more complex indicator for measuring researcher independence on the individual level, based on co-authorship network and topical overlaps. Their measure uses only information about publications’ characteristics (authorship and reference list) regardless of their impacts, unlike other alternative measures that focus solely on the impact of the independent work in terms of received citations (Dey, et al., 2021, @ALISNovelMetric).

Since the explicit goal of many funding instruments that are being used by research funders all over the world is to help early career researchers become independent, we decided to use the newly created indicators to assess the extent, to which such programmes influence researchers' independence.

## Intro 2: Perspective of “What are the effects of competitive grant funding? And looking at independence as one of such effects”

///Primer of grant-making context:

Grant funding mechanisms have previously been shown to stimulate productivity in terms of the number of publications, impact in terms of citations, and even scientific novelty (@wang2018, or as some authors argue rather interdisciplinarity: @fontana2020). In terms of researchers’ careers, obtaining grant funding was connected to significantly higher chances of career progression and obtaining a professorship (@bloch2014).

///Though there is still some disagreement about the direction and magnitude of some of these effects, c.f. @@glaserIndependenceResearchReview2022 on novelty. --> here I can potentially go more in-depth into the critiques of grant funding mechanisms? E.g. spell out the critiques, such that they are strengthening the Matthews effect, they reduce novelty, etc....

However, there are still several components of research excellence (see @vanarensbergenDifferentViewsScholarly2014; @moherAssessingScientistsHiring2018) that have not been investigated yet, with one having specifically important effects on researchers’ carers and development: their independence. Researchers’ independence from their PhD advisor is one of the criteria often used for assessing researchers for hiring, tenure and promotion (@vanarensbergenDifferentViewsScholarly2014; @moherAssessingScientistsHiring2018). The promise of independence is that researchers will “challenge the existing common views and open up new lines of research” (@vandenbesselaarMeasuringResearcherIndependence2019, p.2).

An independent investigator is then defined by the US National Academies of Science (NAS) as “one who enjoys independence of thought—the freedom to define the problem of interest and to choose or develop the best strategies and approaches to address that problem–especially during the earlier career phase” (@vandenbesselaarMeasuringResearcherIndependence2019, p.4). Such a definition does not necessarily put independence in relation to other actors (e.g. one’s PhD advisor), but only in relation to one’s own freedom and resources. However other institutions, such as European Research Council via their Starting Grants scheme (@neufeldPeerReviewbasedSelection2013) think of independence in a relational manner, in terms of having one (or a few) publications without the former PhD supervisor being co-author; being PI on a grant; or being lab director or research leader.

Based on some early empirical studies @rojkoScientificPerformanceResearch2022, it seems that researchers in disciplines such as philosophy, psychology; religion, theology; social sciences; the arts, recreation, entertainment, sport; language, linguistics, literature; and geography, biography, history experience significantly higher levels of independence on their PhD advisor than researchers in disciplines such as science and knowledge, organization, computer science, information, documentation, librarianship, institutions, publications; mathematics, natural sciences; and applied sciences, medicine, technology. 
There is also some preliminary evidence that higher independence modestly correlates with number of citations acquired (@rojkoScientificPerformanceResearch2022), but others argue that even the research systems with very low independence can create a lot of scientific impact, and thus the relationships between independence and impact might not be as clear (@glaserIndependenceResearchReview2022)

In recent decades, there has been a worry that researchers’ indepedence is decreasing, illustrated by increase in age at which US (biomedical) researchers received their first grant (NAS, 2005; @danielsGenerationRiskYoung2015), as well as direct measures of new researchers becoming more dependent on their PhD advisors across almost all disciplines in Slovenia (@rojkoScientificPerformanceResearch2022, table 6 on p.9). Perhaps to tackle this trend, many grant agencies decided to start awarding grants for early career researchers, usually up to 8 years after finishing their PhD. Such programmes have often explicitly stated goals of helping researchers become more independent, set up their own research group and open up new lines of research. The examples of such programmes could be Starting Grants by European Research Council (@StartingGrantERC), Junior project and Junior Star by Czech Science Foundation (2020), Walter N. Benjamin Programme by DFG (@WalterBenjaminProgramme), Innovation Research Incentives Scheme by Netherlands Organization of Scientific Research (NWO) (2021)... something from the US would be also good)

Although researcher independence seems to be important feature that is acknowledged by many actors in the research systems and that research funders try to stimulate, there hasn’t been much effort focused on quantifying this phenomenon. Some studies try to measure independence in a simple way, as a ratio of publications the given researcher produced without their supervisor divided by the total number of publications of the given researchers (e.g. @rojkoScientificPerformanceResearch2022). However, such indicator could be misleading in the world of team science, and also only focuses on collaborations, not tracking whether given researcher become independent in their thinking and research topics. [@vandenbesselaarMeasuringResearcherIndependence2019] therefore proposed more complex indicator for measuring researcher independence on the individual level, based on co-authorship network and topical overlaps. Their measure uses only information about publications’ characteristics (authorship and reference list) regardless of it’s impacts, unlike other alternative measures that focus solely on the impact of the independent work in terms of received citations (Dey, et al., 2021, @ALISNovelMetric).

Since the explicit goal of many funding instruments that are being used by research funders all over the world is to help early career researchers become independent, we used the newly created indicators to assess the extent, to which such programmes influence researchers independence.

H1: Researchers, who received funding from programmes aimed at funding early career researchers, will show more independence from their PhD advisors after receiving the funding than researchers who did not receive funding.

H2: Researchers, who received funding from programmes aimed at funding early career researchers, will show more independence from their PhD advisors compared to similar researchers who received other types of funding at similar point in theri career.

## Intro 3: focusing on support for early career researchers (and why is that important topic) and then mention the independence as one of the ways how we can test whether our support mechanisms for early stage researchers actually work

Funding for early career researchers is an important topic in science funding policy discussions as it might have a significant impact on career progression of the new generations of researchers (@wright2017; @pickett2019; @vandenbesselaar2015). At the same time, the age of receiving a first grant has been growing significantly over the past decades (@powell2016, @christian2021). These trends have motivated some to create new proposals for funding agencies and other research funders in order to support early career researchers (@dewinde2021). Some funding agencies decided to tackle this trend by giving advantage to first-awardees and early career researchers in their usual distribution process, while others have created specific programmes designated to funding young researchers soon after their PhD.

///maybe talk about other types of support that early career researchers get? Or maybe have another paragraph describing what does the life of early career researchers usually look like? (--> perhaps mentioning that the time to finish PhD gets longer and longer, people are now expected to do multiple post-docs, etc... --> perhaps borrow some info from Science of Science book Careers section)///

It is expected, that by giving early-stage researchers funding, they will be able to develop more independence and open up new lines of research (@vandenbesselaarMeasuringResearcherIndependence2019, p.2). Younger researchers are already more likely to come up with new topics and ideas, as found by [@packalen2019] by looking at the usage of new words in research papers. However, they might still feel various kinds of pressures to stick with their mentor’s research agendas and lack of opportunities to become more independent. In recent decades, there has been a worry that researchers’ independence is decreasing, illustrated by increase in age at which US (biomedical) researchers received their first grant (NAS, 2005; @danielsGenerationRiskYoung2015), as well as direct measures of new researchers becoming more dependent on their PhD advisors across almost all disciplines in Slovenia (@rojkoScientificPerformanceResearch2022, table 6 on p.9).
///talk about the balance between experience and independence??///

Perhaps to tackle this trend, many grant agencies decided to start awarding grants for early career researchers, usually up to 8 years after finishing their PhD. Such programmes have often explicitly stated goals of helping researchers become more independent, set up their own research group and open up new lines of research. The examples of such programmes could be Starting Grants by European Research Council (@StartingGrantERC), Junior project and Junior Star by Czech Science Foundation (2020), Walter N. Benjamin Programme by DFG (@WalterBenjaminProgramme), Innovation Research Incentives Scheme by Netherlands Organization of Scientific Research (NWO) (2021)... something from the US would be also good)

Although researcher independence seems to be important feature that is acknowledged by many actors in the research systems and that research funders try to stimulate, there hasn’t been much effort focused on quantifying this phenomenon. Some studies try to measure independence in a simple way, as a ratio of publications the given researcher produced without their supervisor divided by the total number of publications of the given researchers (e.g. @rojkoScientificPerformanceResearch2022). However, such indicator could be misleading in the world of team science, and also only focuses on collaborations, not tracking whether given researcher become independent in their thinking and research topics. [@vandenbesselaarMeasuringResearcherIndependence2019] therefore proposed more complex indicator for measuring researcher independence on the individual level, based on co-authorship network and topical overlaps. Their measure uses only information about publications’ characteristics (authorship and reference list) regardless of it’s impacts, unlike other alternative measures that focus solely on the impact of the independent work in terms of received citations (Dey, et al., 2021, @ALISNovelMetric).

Since the explicit goal of many funding instruments that are being used by research funders all over the world is to help early career researchers become independent, we would like to use the newly created indicators to assess the extent, to which such programmes influence researchers' independence.

@@

We calculated how many (Czech) grants each researcher has received prior to the intervention year (i.e. receiving a Junior Grant).

@@

For the construction of the control group, we implemented propensity matching using the package MatchIt (citation?) and matching exactly by the discipline (42 sub-categories of OECD classification) and treatment year, and approximately by number of publications in total, number of publications in WoS/Scopus, career age, interdisciplinary proportion, number of grants received prior to Junior Grant, career age when receiving the first grant and gender. All of these variables were calculated at the intervention year (i.e. year when the given researcher from experimental group received the Junior Grant). The distance was set to "mahalanobis". The matching was done in 1:3 ratio and with replacement. Manual search for supervisors of the matched control group observations was done by ranking the control group observation by distance from the treatment group observation and first searching for the supervisor of the closest control group observation, and if it was not possible to find it, then repeat the same for the second closest, etc... In the end each treatment observation was matched with exactly one control group observation which was closest but at the same time for which it was possible to identify their supervisor.

We have created 2 control groups: one using the matching above and one similar but specifically containing only researchers who received their first grant (other than Junior Grant) around the same career age as researchers in our intervention group. That allowed us to better track the impact of any grant funding vs specifically Junior Grants funding on researchers’ independence.

@@

BTW: there is an overlap between control group observations: `r overlap_control` observations from the unfunded control group also appears in the funded control group

@@

To assess effects of study field on researcher independence, we chose GLS weighted model to account for unequal group sizes and heteroscedastic variances, while keeping the repeated measures design. First, we compared models with and without the field main effect and observed a notable improvement in model fit for the model where main effect of field was included. This suggests that there are statistically significant differences in the RII based on the field of study. Specifically, adding the field of study to the model resulted in a significant increase in the log-likelihood by `r round(anova_result$L.Ratio[2], 2)` units (p < `r round(anova_result$"p-value"[2], 3)`), emphasizing its importance as a predictor. Specifically, compared to the Engineering and Technology field, being in the Medical and Health Sciences field is associated with a decrease in RII by `r  round(coef(model_summary)["fieldMedical and Health Sciences", 1], 2)` units, on average (p = `r round(coef(model_summary)["fieldMedical and Health Sciences", 4], 3)`), while being in the Social Sciences field is associated with an increase in RII by `r round(coef(model_summary)["fieldSocial Sciences", 1], 2)` units (p = `r round(coef(model_summary)["fieldSocial Sciences", 4], 3)`). There are no statistically significant differences between Natural Sciences and Engineering and Technology. Humanities and Agricultural sciences categories were excluded because they had too few observations.
The interaction between treatment and field was not statistically significant, suggesting that the effect of treatment on independence was consistent across disciplines. The three-way interaction among treatment, independence_timing, and field was not statistically significant. This implies that the interaction effect of treatment and independence_timing on independence remained consistent across different disciplines.

>>>>

@@

We calculated how many (Czech) grants each researcher has received prior and 3 years after the intervention year (i.e. receiving a Junior Grant). We did not use this variable for matching, but for filtering out control group researchers pre-matching.

@@

For the construction of the control group, we implemented propensity matching using the package MatchIt (citation?) and matching exactly by the discipline (42 sub-categories of OECD classification) and treatment year, and approximately by number of publications in total, number of publications in WoS/Scopus, career age, interdisciplinary proportion, number of grants received prior to Junior Grant, career age when receiving the first grant and gender. All of these variables were calculated at the intervention year (i.e. year when the given researcher from experimental group received the Junior Grant). The distance was set to "mahalanobis". The matching was done in 1:1 ratio and with replacement.

Using the matching algorithm, we have created 2 control groups: "Unfunded" group did not receive any grants prior to and 3 years after the treatment year. "Funded" group was matched based on career year at which they received their first grant, ensuring that while treatment group researchers received Junior Grant, "funded" researchers received a different grant at the same time. That allowed us to better track the impact of any grant funding vs specifically Junior Grants funding on researchers’ independence.

@@

There is an overlap between control group observations: `r overlap_control` observations from the unfunded control group also appears in the funded control group, showing that they did eventually received grant at the similar time as those from treatment group.

@@

To assess effects of study field on researcher independence, we chose GLS weighted model to account for unequal group sizes and heteroscedastic variances, while keeping the repeated measures design. First, we compared models with and without the field main effect and observed a notable improvement in model fit for the model where main effect of field was included. This suggests that there are statistically significant differences in the RII based on the field of study. Specifically, adding the field of study to the model resulted in a significant increase in the log-likelihood by `r round(anova_result$L.Ratio[2], 2)` units (p < `r round(anova_result$"p-value"[2], 3)`), emphasizing its importance as a predictor. Specifically, compared to the Engineering and Technology field, being in the Medical and Health Sciences field is associated with a decrease in RII by `r  round(coef(model_summary)["fieldMedical and Health Sciences", 1], 2)` units, on average (p = `r round(coef(model_summary)["fieldMedical and Health Sciences", 4], 3)`), while being in the Humanities and Arts is associated with an increase in RII by `r round(coef(model_summary)["fieldHumanities and the Arts", 1], 2)` units (p = `r round(coef(model_summary)["fieldHumanities and the Arts", 4], 3)`). There are no statistically significant differences between Natural Sciences and Engineering and Technology. Humanities and Agricultural sciences categories were excluded because they had too few observations.
The interaction between treatment and field was not statistically significant, suggesting that the effect of treatment on independence was consistent across disciplines. The three-way interaction among treatment, independence_timing, and field was not statistically significant. This implies that the interaction effect of treatment and independence_timing on independence remained consistent across different disciplines.
